Typed Lua: An Optional Type System for Lua

### 2014-03-16 doc/paper/typedlua.tex
final touches for submission to dyla'14

\author{
\alignauthor
André Murbach Maidl\\
  \affaddr{PUC-Rio}\\
  \affaddr{Rio de Janeiro, Brazil}\\
  \email{amaidl@inf.puc-rio.br}
\alignauthor
Fabio Mascarenhas\\
  \affaddr{UFRJ}\\
  \affaddr{Rio de Janeiro, Brazil}\\
  \email{fabiom@dcc.ufrj.br}
\alignauthor
Roberto Ierusalimschy\\
  \affaddr{PUC-Rio}\\
  \affaddr{Rio de Janeiro, Brazil}\\
  \email{roberto@inf.puc-rio.br}
}

@@

\begin{abstract}
\end{abstract}

\category{D.3.3}
         {Programming Languages}
         {Language Constructs and Features}
\category{D.2.3}
         {Programming Languages}
         {Coding Tools and Techniques}
         [Object-oriented programming]

\terms{Languages,Verification}

\keywords{Type Systems, Lua}

@@

The early error detection and better program structure afforded by
static type checking can lead programmers to migrate their code from
a dynamically typed to a statically typed language, once their simple
scripts become complex programs~\citep{tobin-hochstadt2006ims}.
As this migration involves languages with different syntax and
semantics, it requires a complete rewrite of existing programs intead
of incremental evolution from dynamic to static types.

@@

In this work we outline the design of Typed Lua:
an optional type system for Lua that is complex enough to
preserve some of the idioms that Lua programmers are already used to,
while adding new constructs that help programmers structure Lua
programs.

@@

Lua values can have one of eight {\em tags}: {\em nil}, {\em boolean},
{\em number}, {\em string}, {\em function}, {\em table}, {\em userdata},
and {\em thread}. We will see how Typed Lua assigns types to
values of the first five in this section.

@@

Typed Lua allows optional type annotations in variable function
declarations. It assigns the dynamic type to any parameter that
does not have a type annotation, but assigns more precise
types to unnanotated variables, based on the type of the
expression that gives the initial value of the variable.

@@

The compiler assigns the type $\Number$ to the local variable \texttt{x},
and this example compiles without warnings. Typed Lua allows
programmers to combine annotated code with
unannotated code, as we show in the following example:
\begin{verbatim}
    local function abs(n: number)
      if n < 0 then
        return -n
      else
        return n
      end
    end

    local function distance(x, y)
      return abs(x - y)
    end
\end{verbatim}

@@

Lua has first-class functions, but they have some peculiarities. First,
the number of arguments passed to a function does not need to
match the function's arity; Lua silently drops extra arguments after
evaluating them, or passes {\tt nil} in place of any missing arguments.
Second, functions can return any number of values, and the number
of values returned may not be statically known. Third, Lua also has
multiple assignment, and the semantics of argument passing match
those of multiple assignment (or vice-versa); calling a function is like
doing a multiple assigment where the left side is the parameter list
and the right side is the argument list.

@@

\begin{verbatim}
    local function multiple()
      return 2, "foo"
    end

    local function sum(x: number, y: number)
      return x + y
    end

    local x, y, z = multiple(), multiple()
    print(sum(multiple(), multiple())
\end{verbatim}

@@

Functions that overload their return types to signal the ocurrence
of errors are another common Lua idiom. In this idiom, a function
returns its normal set of return values in case of success but,
if anything fails, returns {\tt nil} as the first value, followed
by an error message or other data describing the error, as in the
following example:

@@

Notice that we could also discriminate {\tt r} using
{\tt type(r) == "number"} as our predicate, with the same
result. The first form is more succint, and more idiomatic.
We can also use projection types to write overloaded functions
where the type of a parameter depends on the type of another
parameter.

@@

\begin{verbatim}
    local t: { string: number? } = { foo = 1 }
    local x: number = t.foo         -- error
    local y: number = t.bar or 0    -- y gets 0
    local z: number? = t["bar"]     -- z gets nil
\end{verbatim}

@@

\begin{verbatim}
    local function greet(person: Person)
      return "Hello, " .. person.firstname ..
             " " .. person.lastname
    end

    local user1 = { firstname = "Lewis",
                    middlename = "Allan",
                    lastname = "Reed" }
    local user2 = { firstname = "Lou" }
    local user3 = { lastname = "Reed",
                    firstname = "Lou" }
    local user4 = { "Lou", "Reed" }

    print(greeter(user1)) -- Hello, Lewis Reed
    print(greeter(user2)) -- Error
    print(greeter(user3)) -- Hello, Lou Reed
    print(greeter(user4)) -- Error
\end{verbatim}

@@

\begin{verbatim}
    local interface Person
      firstname: string
      middlename: string?
      lastname: string
    end

    local user1: Person = { firstname = "Lewis",
                            middlename = "Allan",
                            lastname = "Reed" }
    local user2: Person = { lastname = "Reed",
                            firstname = "Lou" }
\end{verbatim}

@@

Taken individually, the changes to the type of the two
variables look ok, but aliasing makes one of them unsound.
The location of the change also matters, as the next example
shows:

@@

\begin{verbatim}
    local interface Person
      firstname: string
      middlename: string?
      lastname: string
    end

    local user = {}
    user.firstname = "Lou"
    user.lastname = "Reed"
    local person: Person = user
\end{verbatim}

@@

In the first line of the redefinition of {\tt new},
notice how we can call {\tt Shape}'s constructor inside
the overriden constructor. A limitation of this class
system is that the overriden constructor must be a
subtype of the original constructor, so the type of
{\tt radius} has to be very permissive.

@@

Strongtalk \citep{bracha1993strongtalk,bracha1996strongtalk} is
an optionally-typed version of Smalltalk. It has a
polymorphic structural type system that programmers can use
to annotate Smalltalk programs, but type annotations can be
left out; unnanotated terms are dynamically typed, and can
be cast to any static type. The interaction of the dynamic
type with the rest of the type system is unsound, so
Strongtalk uses the dynamically-checked semantics of
Smalltalk when executing programs, even if the programs
are statically typed.

@@

The main feature of Typed Racket's type system is
{\em ocurrence typing}~\citep{tobin:occur}, where
the type of a variable can be refined through tests
using predicates that check the runtime type of
the variable. As these types of checks are common in
other languages, related systems have
appeared~\citep{guha:flow,winther:flow,pearce:whiley}.

@@

>>>>

\author{
\alignauthor
André Murbach Maidl\\
  \affaddr{PUC-Rio}\\
  \affaddr{Rio de Janeiro, Brazil}\\
  \email{amaidl@inf.puc-rio.br}
\alignauthor
Fabio Mascarenhas\\
  \affaddr{UFRJ}\\
  \affaddr{Rio de Janeiro, Brazil}\\
  \email{mascarenhas@ufrj.br}
\alignauthor
Roberto Ierusalimschy\\
  \affaddr{PUC-Rio}\\
  \affaddr{Rio de Janeiro, Brazil}\\
  \email{roberto@inf.puc-rio.br}
}

@@

\begin{abstract}
Dynamically typed languages trade flexibility and ease of
use for safety, while statically typed languages prioritize
the early detection of bugs, and provide a better framework
for structure large programs. The idea of optional typing
is to combine the two approaches in the same
language: the programmer can begin development with
dynamic types, and migrate to static types as the
program matures. The challenge is designing a type system
that feels natural to the programmer that is used to
programming in a dynamic language.

This paper presents the initial design of Typed Lua, an
optionally-typed extension of the Lua scripting language.
The design of Typed Lua's type system has a novel combination
of features that preserves some of the idioms that Lua
programmers are used to, while bringing static type
safety to them.
\end{abstract}

\keywords{Lua programming language, type systems, optional typing, gradual typing}

@@

The early error detection and better program structure afforded by
static type checking can lead programmers to migrate their code from
a dynamically typed to a statically typed language, once their simple
scripts become complex programs~\citep{tobin-hochstadt2006ims}.
As this migration involves languages with different syntax and
semantics, it requires a complete rewrite of existing programs instead
of incremental evolution from dynamic to static types.

@@

This paper presents the initial design of Typed Lua:
an optional type system for Lua that is complex enough to
preserve some of the idioms that Lua programmers are already
used to, while adding new constructs that help programmers
structure Lua programs.

@@

Lua values can have one of eight {\em tags}: {\em nil}, {\em boolean},
{\em number}, {\em string}, {\em function}, {\em table}, {\em userdata},
and {\em thread}. In this section, we will see how Typed Lua assigns types to
values of the first five.

@@

Typed Lua allows optional type annotations in variable function
declarations. It assigns the dynamic type to any parameter that
does not have a type annotation, but assigns more precise
types to unannotated variables, based on the type of the
expression that gives the initial value of the variable.

@@

The compiler assigns the type $\Number$ to the local variable \texttt{x},
and this example compiles without warnings. Typed Lua allows
programmers to combine annotated code with
unannotated code, as we show in the following example:
\begin{verbatim}
    local function abs(n: number)
      if n < 0 then
        return -n
      else
        return n
      end
    end
    local function distance(x, y)
      return abs(x - y)
    end
\end{verbatim}

@@

Lua has first-class functions, but they have some peculiarities. First,
the number of arguments passed to a function does not need to
match the function's arity; Lua silently drops extra arguments after
evaluating them, or passes {\tt nil} in place of any missing arguments.
Second, functions can return any number of values, and the number
of values returned may not be statically known. Third, Lua also has
multiple assignment, and the semantics of argument passing match
those of multiple assignment (or vice-versa); calling a function is like
doing a multiple assignment where the left side is the parameter list
and the right side is the argument list.

@@

\begin{verbatim}
    local function multiple()
      return 2, "foo"
    end
    local function sum(x: number, y: number)
      return x + y
    end
    local x, y, z = multiple(), multiple()
    print(sum(multiple(), multiple())
\end{verbatim}

@@

Functions that overload their return types to signal the occurrence
of errors are another common Lua idiom. In this idiom, a function
returns its normal set of return values in case of success but,
if anything fails, returns {\tt nil} as the first value, followed
by an error message or other data describing the error, as in the
following example:

@@

Notice that we could also discriminate {\tt r} using
{\tt type(r) == "number"} as our predicate, with the same
result. The first form is more succinct, and more idiomatic.
We can also use projection types to write overloaded functions
where the type of a parameter depends on the type of another
parameter.

@@

\begin{verbatim}
    local t: { string: number? } = { foo = 1 }
    local x: number = t.foo       -- compile error
    local y: number = t.bar or 0  -- y gets 0
    local z: number? = t["bar"]   -- z gets nil
\end{verbatim}

@@

\begin{verbatim}
    local function greet(person: Person)
      return "Hello, " .. person.firstname ..
             " " .. person.lastname
    end
    local user1 = { firstname = "Lewis",
                    middlename = "Allan",
                    lastname = "Reed" }
    local user2 = { firstname = "Lou" }
    local user3 = { lastname = "Reed",
                    firstname = "Lou" }
    local user4 = { "Lou", "Reed" }
    print(greeter(user1)) -- Hello, Lewis Reed
    print(greeter(user2)) -- Error
    print(greeter(user3)) -- Hello, Lou Reed
    print(greeter(user4)) -- Error
\end{verbatim}

@@

\begin{verbatim}
    local interface Person
      firstname: string
      middlename: string?
      lastname: string
    end
    local user1: Person = { firstname = "Lewis",
                            middlename = "Allan",
                            lastname = "Reed" }
    local user2: Person = { lastname = "Reed",
                            firstname = "Lou" }
\end{verbatim}

@@

Taken individually, the changes to the type of the two
variables look fine, but aliasing makes one of them unsound.
The location of the change also matters, as the next example
shows:

@@

\begin{verbatim}
    local interface Person
      firstname: string
      middlename: string?
      lastname: string
    end
    local user = {}
    user.firstname = "Lou"
    user.lastname = "Reed"
    local person: Person = user
\end{verbatim}

@@

In the first line of the redefinition of {\tt new},
notice how we can call {\tt Shape}'s constructor inside
the overridden constructor. A limitation of this class
system is that the overridden constructor must be a
subtype of the original constructor, so the type of
{\tt radius} has to be very permissive.

@@

Strongtalk \citep{bracha1993strongtalk,bracha1996strongtalk} is
an optionally-typed version of Smalltalk. It has a
polymorphic structural type system that programmers can use
to annotate Smalltalk programs, but type annotations can be
left out; unannotated terms are dynamically typed, and can
be cast to any static type. The interaction of the dynamic
type with the rest of the type system is unsound, so
Strongtalk uses the dynamically-checked semantics of
Smalltalk when executing programs, even if the programs
are statically typed.

@@

The main feature of Typed Racket's type system is
{\em occurrence typing}~\citep{tobin:occur}, where
the type of a variable can be refined through tests
using predicates that check the runtime type of
the variable. As these types of checks are common in
other languages, related systems have
appeared~\citep{guha:flow,winther:flow,pearce:whiley}.

@@

This paper presented the initial design of Typed Lua,
an optional type system for the Lua scripting language.
While Lua shares several characteristics with other
dynamic languages such as JavaScript, so Typed Lua's
type system has several parts in common with optional
and gradual type systems for dynamic languages,
Lua also has some uncommon features.

These language features demanded features in the type
system that are not present in the type systems of these
other languages: functions with flexible arity and
their interaction with multiple assignment, functions
that are overloaded on the {\em number} of values
they return, and how to support the idiomatic way of
handling their return values, and incremental evolution
of record and object types.

Typed Lua is a work in progress: while the design covers
most of the Lua language as it is actually used by Lua
programmers, there are missing parts. A major omission
of the current version is the lack of polymorphic function
and table types, which will be present in the next version.
The {\tt setmetatable} primitive that we briefly saw in
Section~6 has several other uses that are currently
outside of the scope of the type system: operator
overloading, proxies, changing the behavior of built-in
types, multiple inheritance, etc. Finally, Lua has
one-shot delimited continuations~\citep{james2011yield}
in the form of {\em couroutines}~\citep{moura:coro},
which are also currently ignored by the type system.

Unlike some of the other optional type systems,
the type system of Typed Lua does not have deliberately
unsound parts, but we still do not have proofs that the
novel parts of our type system are sound.

Finally, while we our survey of Lua modules gives some
confidence that the design of Typed Lua is useful for
the majority of Lua programs, we have to validate the
design by using Typed Lua to to type a representative 
sample of Lua modules.
